---
title: "Practical Machine Learning Project"
author: "Camilo Caudillo"
date: "Sunday, March 22, 2015"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset). 

## Data  


The training data for this project are available here: 

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

The test data are available here: 

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

## What you should submit  

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.  

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).  
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.  


```{r getting the data}

# library(httr)
# library(httpuv)
# if(!file.exists("./data")){dir.create("./data")}
# fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# download.file(fileUrl,destfile="./data/pml-training.csv")
# fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# download.file(fileUrl,destfile="./data/pml-testing.csv")


```

## Exploratory data analysis

The training data contains 19622 observations of 160 variables. The response variable is Classe a factor with 5 levels from A to E.

```{r loading packages and data}
library(caret)
set.seed(88388)

training <- read.csv("~/data/pml-training.csv", na.strings=c("NA","#DIV/0!",""))
dim(training)

p <- qplot(factor(classe), data=training, geom="bar")
p + ggtitle("Classe frequency in all the data set")

dev.copy(png, file = "Plot_1.png", width = 1000, height = 800 ) # Saving Plot to local directory. then copy it to the figures folder in the local repository 
dev.off()


```

The response variable is quite balanced in the dataset, the most frecuent class is A, followed by B, E, C and D.

```{r, echo = FALSE}
str(training)
````
## Cleaning the data
Using the `str` command suggest that there are many variables with missing values, we will select only the variables with less than 60% NAs.  
We also have to get rid of the first variables, Unique ID, User, dates, times etc...

```{r}
train=training[,colSums(is.na(training))<(nrow(training)*0.6)]
dim(train)
str(train)
train=subset(train, select=-c(X,user_name,raw_timestamp_part_1,raw_timestamp_part_2,cvtd_timestamp,new_window,num_window))
dim(train)
str(train)
````
We have now only 52 explanatory variables in numeric and integer format.  

# Training the model.
We will create the training and testing partitions, take samples of 1000 rows for train a random forest model, we expect this is the best method.

````{r}
inTrain <- createDataPartition(train$classe, p=0.7,list=FALSE) 
length(inTrain)

inTrain=sample(inTrain,size=1000,replace=FALSE) #subset for testing
training <- train[inTrain,]
testing <- train[-inTrain,]
testing=testing[sample(nrow(testing), 1000),] #random rows
modelrf=train(classe~.,data=training,method="rf")
modelrf
````

The random forest model accuracy is pretty good 95% CI: (0.8884, 0.9252), and the kappa index is acceptable 0.88  

##Testing error estimation  
We need now to get the testing error.  

````{r}

predictions=predict(modelrf,newdata=testing)
accuracy=sum(predictions == testing$classe)/length(testing$classe) #or:
confusionMatrix(predictions,testing$classe)
````
In the testing set the random forest model gets 90.8% accuracy.
 
Maybe there is information redundancy, we can reduce the predictors to the first 20 variables in importance to the model.

```{r}
varImp(modelrf)

inTrain <- createDataPartition(train$classe, p=0.7,list=FALSE) 
length(inTrain)
training <- train[inTrain,]
testing <- train[-inTrain,]
training=subset(training,select=c(roll_belt,pitch_forearm,magnet_dumbbell_z,magnet_dumbbell_y,yaw_belt,pitch_belt,roll_forearm,magnet_dumbbell_x,roll_dumbbell,magnet_belt_z,accel_forearm_x,accel_dumbbell_y,gyros_dumbbell_y,roll_arm,magnet_belt_x,accel_belt_z,magnet_belt_y,accel_dumbbell_z,magnet_forearm_x ,magnet_forearm_x,total_accel_dumbbell,classe))
testing=subset(testing,select=c(roll_belt,pitch_forearm,magnet_dumbbell_z,magnet_dumbbell_y,yaw_belt,pitch_belt,roll_forearm,magnet_dumbbell_x,roll_dumbbell,magnet_belt_z,accel_forearm_x,accel_dumbbell_y,gyros_dumbbell_y,roll_arm,magnet_belt_x,accel_belt_z,magnet_belt_y,accel_dumbbell_z,magnet_forearm_x ,magnet_forearm_x,total_accel_dumbbell,classe))
dim(training)

modelrf=train(classe~.,data=training,method="rf")
modelrf
````

The accuracy with the entire dataset is about 98%.  

##Training all data with repeated crossvalidation
To get the out-of-sample error estimate we will train a random forest on the whole training data using cross-validation, specifically 10 fold repeated cross validation.

```{r}

fitControl <- trainControl(method = "repeatedcv", 
                           number = 10, repeats = 10)
model=train(classe~.,data=training,method="rf",trControl=fitControl)

model

````
We get both accuracy and kappa values of 0.98 in the repeated crossvalidation



##Final model applied to the test data


```{r}

model=train(classe~.,data=training,method="rf")
predictions=predict(model,newdata=testing)
accuracy=sum(predictions == testing$classe)/length(testing$classe)
accuracy
confusionMatrix(predictions,testing$classe)

````

In the test data the accuracy IC is (0.99, 0.9946), andthe kappa index is ranging from 0.98688 in class B to 0.9970 in class A.


##Predicting the Classe variable in the testing data

We now need to execute the predictions on the blind data, first we select only the variables in the model.

````{r}

testing <- read.csv("~/data/pml-testing.csv", header=TRUE, na.strings=c("NA","#DIV/0!",""))

testing=subset(testing,select=c(roll_belt,pitch_forearm,magnet_dumbbell_z,magnet_dumbbell_y,yaw_belt,pitch_belt,roll_forearm,magnet_dumbbell_x,roll_dumbbell,magnet_belt_z,accel_forearm_x,accel_dumbbell_y,gyros_dumbbell_y,roll_arm,magnet_belt_x,accel_belt_z,magnet_belt_y,accel_dumbbell_z,magnet_forearm_x ,magnet_forearm_x,total_accel_dumbbell))
dim(testing)
predictTest=predict(model,newdata=testing)
length(predictTest)

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(predictTest)
````
